Type Refinements for the Working Class

### 2015-12-22 paper.tex
Add dependent function types

>>>>

\paragraph{Functional Typehood}
\newcommand\PTyFamNil{\vDash^\textit{type}_{\mathtt{nil}}}
\newcommand\PTyFamSnoc{\vDash^\textit{type}_{\mathtt{snoc}}}

Let us define functional typehood
$\EqTypeFam{\Upsilon}{\Psi}{\phi}{\psi}{\tau}$, presupposing
$\FEqRefines{\Upsilon}{\Psi}{\phi}{\psi}{\tau}$. As usual, we
give our meaning explanation by induction on the evidence for
the validity of the refined context $\Psi$.

\paragraph{Case} $\DefRefinesCtxNil{\Upsilon}$.
\[
  \infer[\PTyFamNil]{
    \EqTypeFam{\Upsilon}{\cdot}{\phi}{\psi}{\tau}
  }{
    \IsType{\Upsilon}{\phi}{\tau}
  }
\]
Note that this case is not trivial, because it is only defined in
case the presupposition $\FEqRefines{\Upsilon}{\Psi}{\phi}{\psi}{\tau}$
obtains. It suffices to require only that $\phi$ be a type, since because
typehood respects equality of refinement.

\begin{quote}
  To know $\EqTypeFam{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}$ is to know,
  for any renaming
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
  and closed terms $\Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}$, that from
  $\FRMemEq{\Upsilon'}{\Rename{\rho}{\Psi}}{M_0}{M_1}{\Rename{\rho}{\chi}}$, you can conclude
  $\EqTypeFam{\Upsilon'}{\Rename{\rho}{\Psi}}{\Subst{M_0}{x}{\Rename{\rho}{\phi}}}{\Subst{M_1}{x}{\Rename{\rho}{\psi}}}{\tau}$.
  In other words:
  %
  \[
    \infer[\PTyFamSnoc]{
      \EqTypeFam{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}
    }{
      \begin{array}{l}
        \forall
        \begin{tikzcd}[cramped, ampersand replacement = \&]
          \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] \&\IMode{\Upsilon'}
        \end{tikzcd}
        .\ \forall \Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}.\\
        \left(\FRMemEq{\Upsilon'}{\Rename{\rho}{\Psi}}{M_0}{M_1}{\Rename{\rho}{\chi}}\right)\\
        \quad\implies
        \left(\EqTypeFam{\Upsilon'}{\Rename{\rho}{\Psi}}{\Subst{M_0}{x}{\Rename{\rho}{\phi}}}{\Subst{M_1}{x}{\Rename{\rho}{\psi}}}{\tau}\right)
      \end{array}
    }
  \]
\end{quote}

As with refinements, we will write $\IsTypeFam{\Upsilon}{\Psi}{\phi}{\tau}$
to mean $\EqTypeFam{\Upsilon}{\Psi}{\phi}{\phi}{\tau}$.

\subsection{Example: Dependent Function Types}
\newcommand\SortArr[2]{#1\rightharpoonup #2}
\newcommand\OpLambda[2]{\mathtt{lam}_{#1}^{#2}}
\newcommand\OpAp[2]{\mathtt{ap}_{#1}^{#2}}

Fix an extensional lcs such that its signature contains the following:
\[
  \infer{
    \IsSort{\SortArr{\sigma}{\tau}}
  }{
    \IsSort{\sigma} &
    \IsSort{\tau}
  }
  \qquad
  \infer{
    \IsOperator{\Upsilon}{
      \OpAp{\sigma}{\tau}
    }{
      \MkArity{
        \MkValence{}{}{\SortArr{\sigma}{\tau}};
        \MkValence{}{}{\sigma}
      }{\tau}
    }
  }{
    \IsSort{\sigma} &
    \IsSort{\tau}
  }
\]

The sort $\SortArr{\sigma}{\tau}$ is intended to classify the (syntax of)
partial operations, and $\OpAp{\sigma}{\tau}$ is the application operator. Now,
we can define a refinement on this sort for dependent functions, and then prove
that it is a type. We intend to define a refinement $\prod_{(x:\phi)}\psi$
such that the following rule holds:
\[
  \infer{
    \Refines{\Upsilon}{\prod_{(x:\phi)}\psi}{\SortArr{\sigma}{\tau}}
  }{
    \Refines{\Upsilon}{\phi}{\sigma} &
    \FRefines{\Upsilon}{x:\phi}{\psi}{\tau}
  }
\]

For any renaming
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}%
, the meaning of $\prod_{(x:\phi)}\psi$ at
$\Member{M, N}{\Exprs_{\SortArr{\sigma}{\tau}}(\Upsilon')}$ is given by
the following rule:
\[
  \infer={
    \RMemEq{\Upsilon'}{M}{N}{
      \prod_{(x:\phi)}\psi
    }
  }{
    \FRMemEq{\Upsilon'}{x:\phi}{
      \App{\OpAp{\sigma}{\tau}}{M; x}
    }{
      \App{\OpAp{\sigma}{\tau}}{N; x}
    }{
      \psi
    }
  }
\]

Note that we have not required an introduction form such as
$\OpLambda{\sigma}{\tau}$, and have instead opted for a negative
definition of the dependent function refinement. This is strictly more
open-ended than the positive definition, because it does not fix in
advance a syntax for the canonical formation of functions.

\begin{thm}
  The dependent function refinement
  $\Refines{\Upsilon}{\prod_{(x:\phi)}\psi}{\SortArr{\sigma}{\tau}}$ is a
  type assuming $\IsType{\Upsilon}{\phi}{\sigma}$ and
  $\IsTypeFam{\Upsilon}{x:\phi}{\psi}{\tau}$.
\end{thm}
\begin{proof}
  Fix
    \begin{tikzcd}[cramped]
      \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
    \end{tikzcd}%
  , and $\Member{L,M,N}{\Exprs_{\SortArr{\sigma}{\tau}}(\Upsilon')}$ such that
  \begin{enumerate}
    \item $\PBinRel{\sim_{\SortArr{\sigma}{\tau}}}{\Upsilon'}{L}{M}$
    \item $\FRMemEq{\Upsilon'}{x:\Rename{\rho}{\phi}}{\App{\OpAp{\sigma}{\tau}}{M;x}}{\App{\OpAp{\sigma}{\tau}}{N;x}}{\Rename{\rho}{\psi}}$
  \end{enumerate}
%
  It suffices to show that
  $\FRMemEq{\Upsilon'}{x:\Rename{\rho}{\phi}}{\App{\OpAp{\sigma}{\tau}}{L;x}}{\App{\OpAp{\sigma}{\tau}}{N;x}}{\Rename{\rho}{\psi}}$.
%
  Now fix
    \begin{tikzcd}[cramped]
      \IMode{\Upsilon'}\arrow[r,hook,"\IMode{\rho'}"] &\IMode{\Upsilon''}
    \end{tikzcd}%
  , and $\Member{X,Y}{\Exprs_\sigma(\Upsilon'')}$ such that
  $\RMemEq{\Upsilon''}{X}{Y}{\Rename{(\rho'\circ\rho)}{\phi}}$.
  Then, we have to show that
  $\RMemEq{\Upsilon''}{%
    \App{\OpAp{\sigma}{\tau}}{
      \Rename{\rho'}{L};
      X
    }
  }{
    \App{\OpAp{\sigma}{\tau}}{
      \Rename{\rho'}{N};
      Y
    }
  }{\Subst{X}{x}{\Rename{(\rho'\circ\rho)}{\psi}}}$.

  Because $\psi$ is a functional family of types,
  $\Subst{X}{x}{\Rename{(\rho'\circ\rho)}{\psi}}$ must respect
  computational equivalence, so by drawing hypothesis (1) along $\rho'$, we
  can rewrite hypothesis (2) into the goal.
\end{proof}
